PyTracer: Automatically profiling numerical instabilities in Python

### 2021-11-04 paper.tex
Tristan review

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computational codes. Despite the recent large numbers of tools dedicated to evaluating codes' numerical stability, analyzing a complete program remains a handy task for the user. To address this issue, we propose \pytracer, an automatic profiler for analyzing numerical instabilities of Python applications. \pytracer automatically instruments Python code to produce numerical traces that can be visualized through a Dash server, allowing an interactive visualization of numerical stability.
\pytracer is designed to work with any random noise model that simulates numerical instabilities, and we demonstrate its usage with the Monte Carlo Arithmetic through the fuzzy ecosystem.
Finally, we demonstrate that pytracer can handle intensively used Python scientific libraries such as NumPy, SciPy, and Scikit-learn on a representative set of tests, at low cost with an average slowdown $\times 3.5$.
\end{abstract}

@@

The scientific Python ecosystem is a central component of scientific
analyses due to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Diverse research groups built numerous domain-specific data analysis pipelines from this ecosystem, leveraging Python's simplicity and expressivity.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by data noise, software and hardware updates, or parallelization lead to substantial deviations in final results and potentially different scientific conclusions, threatening the reliability of computational analyses. Numerical stability analyses need to be conducted systematically to address this issue. However, no practical tool currently exists to conduct such analyses on Python programs.

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces, enabling scaling to large programs without manual intervention. In contrast, theoretical approaches based on condition numbers or backward error analysis require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques such as Frama-C~\cite{cuoq2012frama}, Gappa~\cite{de2010certifying} or Flocq~\cite{boldo2011flocq} hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

PyTracer estimates the number of significant digits of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) tracing floating-point computations along program executions, (2) generating numerically-perturbed executions, and (3) visualizing stability evaluations. PyTracer addresses these requirements with (1) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (2) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

Through stability evaluations of 
SciPy, scikit-learn, and PyAFQ~\cite{kruper2021evaluating} --- an analysis toolbox for diffusion magnetic resonance imaging --- we demonstrate PyTracer as a practical solution to review the numerical stability of extensive code bases. The remainder of this manuscript describes related approaches, presents the design of PyTracer, and demonstrates its potential in various use cases.

@@

Several tools have been and continue to be developed to assess numerical quality. Those tools can be divided into two main categories depending on whether they adopt a static or a dynamic approach.
Static analysis produces theoretically valid error bounds, while dynamic analysis is generally more scalable, although it only provides experimental estimations.
\pytracer adopts the dynamic approach due to its scalability to large existing code bases. Dynamic approaches typically trace floating-point computations, detect instabilities in traces, and visualize summary statistics about instabilities. The remainder of this section reviews existing approaches for each of these steps.

@@

The source-to-source technique requires a rewriting of the application to modify floating-point types. Among frameworks that adopted this technique,
CADNA~\cite{jezequel2008cadna} is a library for C, C++ and Fortran implementing the CESTAC~\cite{vignes1993stochastic} stochastic arithmetic.
Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
Finally, the work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without  stochastic arithmetic.

The compiler-based approach uses a compiler to analyze or replace floating-point expressions with other models automatically. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability using a source-to-source transformation at the Abstract Syntax Tree (AST)  level to rewrite parts of code with higher precision. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel}, is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

Dynamic Binary Instrumentation (DBI) operates on an executable directly, without the need for recompilation or manual changes, and is thus
transparent to the programming language used. 
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that replaces on-the-fly
floating-point operations by their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical inaccuracies by using shadow computations with higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to detect precision losses by comparison with results obtained with the high-precision 
MPFR library~\cite{fousse2007mpfr}. It is combined with symbolic computation to backtrack the root of the error.
We can note that all methods based on a high-precision oracle suppose that computing with a large number of bits is sufficient to obtain an accurate result, which is not always true, see, for example, the famous Muller's sequence ~\cite{chatelain2018veritracer}.

Being agnostic to the programming language is a serious advantage of the DBI method compared to source-to-source and compiler-bases methods. However, working at a binary level makes it difficult to access high-level information needed to debug and understand the source code logic. Conversely, the source-to-source approach provides a fine-grained control on the code being analyzed, bu itt lacks scalability, as rewriting large codes can be a tedious task. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source libraries.

To the best of our knowledge, there is currently no tool dedicated to the numerical analysis of Python code. Existing tools for tracing Python code are dedicated to performance profiling, for time (cProfile) 
or memory consumption (memprofile). 
Anteater~\cite{faust2019anteater} is a Python tracer tool to debug Python applications. 
It performs source transformations at the AST level but only deals with native numeric Python types.
Moreover, the user needs to manually tag each variable to trace. Finally, according to the authors, Anteater does not scale to large traces.
cProfile, memprofile, and Anteater use Python decorators as the main instrumentation technique, a Python mechanism to instrument a function by adding a line over a function declaration.
While this method is appropriate when targeting specific functions, it is not feasible for large code bases where the location of potentially unstable code sections is unknown.

\label{sec:detecting-instabilities}
\subsection{Detecting numerical instabilities}

Three main approaches exist to detect numerical instabilities once floating-point computations are traced: stochastic arithmetic, uncertainty or sensitivity analysis, and random seed analysis. The main techniques for each approach are reviewed below.

@@

\label{sec:mca}
\subsubsection{Stochastic Arithmetic}

@@

Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors 
% \tristan{only round-off errors or all numerical errors?} 
as random variables and to characterize them statistically.

The first published complete stochastic arithmetic framework was CESTAC~\cite{vignes1993stochastic}, where each floating-point computation is performed multiple times with different rounding modes: round-up or round-down. From these stochastic samples, one can then derive the number of significant digits in the computed result. Discrete Stochastic Arithmetic~\cite{vignes2004discrete} extends CESTAC to comparison operators implemented in the CADNA library.

Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} uses the same principle but introduces two differences:
(i) a virtual precision parameter, allowing to simulate reduced working precision,  and (ii) different perturbation modes, to introduce perturbations
on function inputs or output.
MCA simulates round-off errors by introducing random perturbations in floating-point computations. It replaces each floating-point number with its stochastic counterpart:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a uniform variable.
Virtual precision allows for the simulation of reduced working precision.
MCA can be applied in three modes: Random Rounding (RR), Precision Bounding (PB), and full MCA, which respectively perturb either the output, the inputs, or both with the $inexact$ function. While the RR mode is equivalent to stochastic rounding, the PB mode allows identifying catastrophic cancellations.

In stochastic arithmetic, numerical error is measured as the number of significant digits $s$ estimated among the sampled values. A common formula to determine this number from MCA samples is the one reported in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  Recent variations of this formula that also provide confidence intervals are discussed in~\cite{sohier2018confidence}.

\subsubsection{Uncertainty, sensitivity, and random seed analysis}

Uncertainty analysis (UA) is a method to estimate the output distribution given an input distribution.
Sensitivity analysis (SA) is a variation of UA that measures the local response to a small perturbation~\cite{loucks2017water}. 
Several approaches for UA and SA exist, among them differential analysis, response surface methodology, Monte Carlo analysis, and variance decomposition (see~\cite{helton2006survey} for a review). In particular, Monte Carlo analysis is a sampling-based technique that samples inputs space to estimate output response. Monte Carlo analysis shares similarities with Monte Carlo Arithmetic. For instance, the PRECISE framework~\cite{chaitin1996lectures} shares with MCA the idea of varying virtual precision to estimate output sensitivity.
Furthermore, the quality of a Random Number Generator is a crucial characteristic when dealing with stochastic model and differ from cryptography requirements. For numerical simulation, a good RNG provides a tradeoff between performance, period, correlation, or uniformity~\cite{james2020review}. On the other hand, poor quality RNGs may have a substantial impact on results~\cite{click2011quality}. First, measuring variation induced by seed is a good indicator of program stability that we can compare against variability induced by floating-point arithmetic.

@@

Visualization covers a wide range of numerical simulation domains, from general-purpose numerical visualization with tools
such as VTK~\cite{schroeder2000visualizing}, to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
We review here tools dedicated to the analysis of numerical variability induced by the floating-point model. Despite the large number of existing tools  to locate unstable source code sections, few of them allow the user to have a global view of the numerical accuracy of their computations or compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer of numerical stability over time for C, C++, and Fortran codes. It is an extension of Verificarlo~\cite{verificarlo} that offers contextual information about variables traced to allow the user to identify instabilities quickly.
However, it does not offer a dynamic interface and does not include information about high-level structures such as \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} visualizes the effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. However, Anteater provides an insightful reflection 
% \tristan{It's a bit vague. Do you simply mean: ``Anteater's GUI inspired the design of \pytracer's interface"?} 
around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

% \subsection{Pytracer workflow}

The PyTracer workflow follows a three-step approach (Figure~\ref{fig:workflow}). First, \pytracer instruments applications using ``monkey patching", a technique to modify Python programs without editing their source code. The instrumentation replaces all the functions with a wrapper that saves their arguments and returned value in a trace file using ``pickle", Python's object serialization mechanism. \pytracer then aggregates trace files obtained from multiple perturbed executions, computing summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the original application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable; in this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches, which is scalable and useful in read-only environments such as Singularity containers or HPC clusters.

@@

Python loads modules through an import mechanism triggered by the \texttt{import} keyword, using two objects: \texttt{finders} and \texttt{loaders}.
The \texttt{finder} is responsible for finding the package (or 
namespace\footnote{Python distinguishes between \texttt{regular} and \texttt{namespace} packages.
A regular package is a directory that contains a \_\_init\_\_.py file and potentially subdirectories (sub-packages) 
that contain themselves a \texttt{\_\_init\_\_.py} file and so on recursively. 
The package hierarchy follows those of the directory. 
Namespace packages introduced in Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not contain an
\_\_init\_\_.py file and allow for flexible directory structures. Hence, parts of the package can be located in zip files, on the network, or in separated directories. There are no functional differences between both types of packages.}) from its fully qualified name, whether it is located on the local storage, such as standard packages installed through the pip package manager, or on a distant server.
\texttt{Finders} do not load the modules; they return a specification (\texttt{spec} object) encapsulating 
information on where to find it and how to load it.
The \texttt{loader} creates and initializes a module object, filling the import-related attributes 
such as \texttt{\_\_spec\_\_}. 
Then, the \texttt{loader} executes the module and populates its namespace. Finally, the module is bound to its import name in the \texttt{sys.modules} dictionary.

Python supports custom \texttt{finder} classes, registered in the \texttt{sys.meta\_path} list.
When Python encounters an \texttt{import} statement, it first looks for a binding in the \texttt{sys.modules}
and then iterates over the \texttt{finder} classes in \texttt{sys.meta\_path} until it finds the module to import. \pytracer adds a custom \texttt{finder} class at the head of the list that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module, then copies it as a new instance of the \texttt{ModuleType} class. It then calls the appropriate instrumentation function for each module attribute depending on its type (function, class, or instances). Then, \pytracer replaces the original module with its instrumented counterpart in the \texttt{sys.modules} map once it has instrument all the attributes. Hence, the calling convention remains unchanged, and the application will transparently call the instrumented module. Finally, Pytracer updates all existing references to the original module, such as the ones in \texttt{\_\_globals\_\_} function attributes that contain references to internal function objects.

@@

For each function, \pytracer's \texttt{loader} class creates a wrapper function, dynamically compiles it with the compile builtin function, and substitutes it for the original module function. This simple technique
cannot be applied to callable class instances, i.e., class instances that have the \texttt{\_\_call\_\_} attribute but are not of the \texttt{FunctionType} type. Indeed, applying the wrapper technique would 1) modify the type of the class instance to \texttt{FunctionType}, which could cause syntactic and semantic bugs, and 2) mask class attributes, leading to \texttt{AttributeError} exceptions.
To overcome this issue, \pytracer overrides the \texttt{\_\_call\_\_} attribute with the wrapper function when possible. When the \texttt{\_\_call\_\_} attribute is not writable,  Pytracer does not instrument the class and returns a warning. 
Listings~\ref{fig:wrapper_creation} and~\ref{fig:generic_wrapper} show \pytracer's wrapper creation function.

@@

    Returns:
        wrapper_code: The code source of the wrapper
    """
    function_id = id(function)
    info = (function_id, function_module, function_qualified_name)
    wrapper_code = f"""
    def {function_wrapper_name}(*args, **kwargs):
        return generic_wrapper({info},*args,**kwargs)"""
    return wrapper_code
\end{lstlisting}
    \caption{Function to create the instrumented version of a function.
    \pytracer uses the identifier of the function instead of the 
    actual function to get rid of aliases and duplicated instrumentations.}
    \label{fig:wrapper_creation}
\end{listing}

@@

\pytracer instruments classes by wrapping their callable attributes as described previously. 
This instrumentation preserves class types, avoiding type mismatches in the instrumented application.
However, some classes, in particular base types and NumPy's universal functions (\texttt{ufuncs}), contain read-only attributes that cannot be instrumented. 
By default, \pytracer returns a warning when it encounters such classes.

\subsubsection{Instrumenting instances}

>>>>

\begin{abstract}
Numerical stability is a crucial requirement of reliable scientific computing. However, despite the pervasiveness of Python in data science, analyzing large Python programs remains challenging due to the lack of scalable numerical analysis tools available for this language. To address this issue, we developed \pytracer, a profiler to quantify numerical instability in Python applications. \pytracer automatically instruments Python code to produce numerical traces and visualize them interactively in a Plotly dashboard.
\pytracer is designed to work with any type of numerical noise model, including Monte-Carlo Arithmetic, random rounding or random data perturbations. We illustrate \pytracer's capabilities by testing the numerical stability of key functions in the SciPy and Scikit-learn Python libraries. \tristan{Add one sentence to summarize the findings} Finally, \pytracer slows down executions by only $3.5 \times$ on average.
\end{abstract}

@@

The scientific Python ecosystem is a central component of scientific
analyses due to its rich offering of data manipulation, array programming,
numerical analysis, and visualization tools. In particular, libraries such
as NumPy~\cite{harris2020array}, SciPy~\cite{virtanen2020scipy}, scikit-learn~\cite{pedregosa2011scikit} or PyTorch~\cite{paszke2019pytorch} are used in hundreds of publications every year, providing a reference set of high-quality open-source core scientific libraries. Researchers have built numerous domain-specific data analysis pipelines from this ecosystem, leveraging Python's simplicity and expressivity.

Numerical stability is a crucial requirement of reliable scientific data
processing. In unstable analyses, small numerical perturbations introduced by data noise, software and hardware updates, or parallelization lead to substantial deviations in final results and potentially different scientific conclusions, threatening the reliability of computational analyses. To address this issue, numerical stability analyses need to be conducted systematically, however, no practical tool currently exists to conduct such analyses for Python programs.

We present PyTracer, a numerical stability analyzer for Python programs.
PyTracer adopts a dynamic approach that evaluates numerical stability using program execution traces and is applicable to large programs without manual intervention. In contrast, theoretical approaches based on condition numbers or backward error analysis require detailed modeling of the program and its numerical implementation. Similarly, static code analysis techniques hardly scale to large codebases, particularly in dynamically-typed languages such as Python.

PyTracer estimates the variability of floating-point variables by combining program traces obtained with different numerical perturbations, which requires (1) generating numerically-perturbed executions, (2) tracing floating-point computations along program executions, and (3) visualizing stability evaluations. PyTracer addresses these requirements with (1) a ``fuzzy" Python interpreter instrumented with Monte-Carlo arithmetic, (2) a dynamic instrumentation of Python functions, modules, and classes through meta-programming, (3) an interactive Plotly dashboard to visualize stability measures throughout program executions.

Through stability evaluations of the
SciPy and scikit-learn libraries, we demonstrate \pytracer as a practical solution to review the numerical stability of extensive code bases. The remainder of this manuscript describes related approaches, presents the design of \pytracer, and demonstrates its potential in various use cases.

@@

Several tools have been developed to assess numerical quality, divided into static and dynamic approaches. Static approaches typically analyze source code while dynamic approaches trace floating-point computations throughout executions. A detailed review of these approaches is presented in~\cite{cherubin2020tools}. 
\pytracer adopts the dynamic approach due to its scalability to large existing code bases. Dynamic approaches typically trace floating-point computations, detect instabilities in traces, and visualize summary statistics about instabilities. The remainder of this section reviews existing approaches for each of these steps.

@@

The source-to-source technique requires a rewriting of the application to modify floating-point types. 
CADNA~\cite{jezequel2008cadna} is a library for C, C++, and Fortran implementing the CESTAC~\cite{vignes1993stochastic} stochastic arithmetic. Shaman~\cite{demeure_phd} is a C++ library that uses a first-order error model to propagate numerical error. 
MCAlib~\cite{frechtling2015mcalib} is a C library that implements the Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} using the MPFR~\cite{fousse2007mpfr} library.
Finally, the work in~\cite{tang2016software} proposed a source-to-source framework for executing targeted code in infinite and fixed precision with and without  stochastic arithmetic.

The compiler-based approach instruments floating-point expressions at compile time. 
Verificarlo~\cite{verificarlo} is a compiler that supports different types of arithmetic instrumentations, including MCA. 
The work in~\cite{bao2013fly} modified the GNU Compiler Collection (GCC) to track local floating-point errors across executions. pLiner~\cite{guo2020pliner} is a root-cause analyzer based on the Clang compiler that detects floating-point operations responsible for result variability using a source-to-source transformation at the Abstract Syntax Tree (AST) level to rewrite parts of code with higher precision. 
PFPSanitizer~\cite{chowdhary2020debugging,chowdhary2021parallel} is a compiler that uses parallel shadow execution to detect numerical issues using higher precision.
FLiT~\cite{sawaya2017flit} is a framework to detect variability induced by compilers and their optimizations.
The work in~\cite{wang2012development} proposes a numerical debugger based on GDB~\cite{stallman1988debugging} for Discrete Stochastic Arithmetic (DSA) on FPGA as an Eclipse plugging. Similarly, Cadtrace~\cite{jezequel2008cadna} and Shaman propose a GDB-based tool to use the CADNA and Shaman libraries with GDB, respectively.

Dynamic Binary Instrumentation (DBI) operates directly on executables, without the need for recompilation or manual changes. Therefore, it is applicable to any programming language.
CraftHPC~\cite{lam2013dynamic} uses DBI to detect catastrophic cancellations at runtime.
Verrou~\cite{fevotte2016verrou} is a Valgrind~\cite{nethercote2007valgrind} based tool that dynamically replaces
floating-point operations with their stochastic arithmetic counterparts. FPDebug~\cite{benz2012dynamic} uses DBI to detect numerical inaccuracies by using shadow computations with higher precision.
Herbgring~\cite{sanchez2017finding} is a Valgrind-based tool to detect
numerical instabilities. It uses a shadow memory to detect precision losses by comparison with results obtained at higher precision. It is combined with symbolic computation to backtrack the root of the error.
We can note that all methods based on a high-precision oracle suppose that computing with a large number of bits is sufficient to obtain an accurate result, which is not always true (see for example, the famous Muller's sequence~\cite{chatelain2018veritracer} \tristan{it looks like your veritracer paper is the first time where the ``famous Muller's sequence" was described, is that so?}).

Being agnostic to the programming language is a serious advantage of the DBI method compared to source-to-source and compiler-bases methods. However, working at a binary level makes it difficult to access high-level information needed to debug and understand the source code logic. Conversely, the source-to-source approach provides a fine-grained control on the analyzed code, but it lacks scalability, as rewriting large codes can be a tedious task. Finally, the compiler-based approach takes advantage of the best of both by being automatic and having access to high-level information. However, like source-to-source, the compiler-based method is not suitable for analyzing closed-source libraries.

To the best of our knowledge, there is currently no tool dedicated to the numerical analysis of Python code. Existing tools for tracing Python code focus on  performance profiling for time (cProfile) 
or memory consumption (memprofile). 
Anteater~\cite{faust2019anteater} is a Python tracer tool to debug Python applications. 
It performs source transformations at the AST level but only deals with native numeric Python types.
Moreover, the user needs to manually tag each variable to trace. Finally, according to the authors, Anteater does not scale to large traces.
cProfile, memprofile, and Anteater use Python decorators as the primary instrumentation technique, a Python mechanism to instrument a function by adding a line over a function declaration.
While this method is appropriate when targeting specific functions, it is not feasible for large codebases where potentially unstable code sections are unknown.

\label{sec:detecting-instabilities}
\subsection{Detecting numerical instabilities with stochastic arithmetic}
\label{sec:mca}

@@

% \subsubsection{Stochastic Arithmetic}

@@

Three main approaches exist to detect numerical instabilities: stochastic arithmetic, uncertainty or sensitivity analysis, and random seed analysis. \tristan{add key references for uq/sa if any} \pytracer focuses on stochastic arithmetic \tristan{due to...}
Stochastic arithmetic leverages randomness to estimate numerical instabilities coming from the use of floating-point representations. The main idea is to treat round-off errors 
% \tristan{only round-off errors or all numerical errors?} 
as random variables and characterize them statistically.

The first published complete stochastic arithmetic framework was CESTAC~\cite{vignes1993stochastic}, where each floating-point computation is performed multiple times with different rounding modes: round-up or round-down. From these stochastic samples, one can then derive the number of significant digits in the computed result. Discrete Stochastic Arithmetic~\cite{vignes2004discrete} extends CESTAC with comparison operators implemented in the CADNA library.

Monte Carlo Arithmetic (MCA)~\cite{parker1997monte} uses the same principle but introduces two differences:
(i) a virtual precision parameter, allowing to simulate reduced working precision,  and (ii) different perturbation modes to introduce perturbations
on function inputs or output.
MCA replaces each floating-point number with its stochastic counterpart:
\[
inexact(x) =  x + \beta^{e_x - t}\xi
\]
where $\beta$ is the number base, $t$ the virtual precision and $\xi \in (-\frac{1}{2},\frac{1}{2})$ is a random uniform variable.
Virtual precision allows for the simulation of reduced working precision.
MCA can be applied in three modes: Random Rounding (RR), Precision Bounding (PB), and full MCA, which perturb either the output, the inputs, or both with the $inexact$ function. While the RR mode is equivalent to stochastic rounding, the PB mode allows identifying catastrophic cancellations.

Stochastic arithmetic uses the number of significant digits $s$ estimated among the sampled values to quantify the numerical errors.
A common formula to determine this number from MCA samples is the one reported in~\cite{parker1997monte}:
\begin{equation}
s = -\log_{\beta}{ \left| \dfrac{\sigma}{\mu} \right|} \label{eq:sig-digits}
\end{equation}
where $\mu$ and $\sigma$ are the sample mean and standard deviation of a variable sampled with MCA.  
Sohier et al.~\cite{sohier2018confidence} recently provided a rigorous generalization of this formula to include confidence intervals.

% \subsubsection{Uncertainty, sensitivity, and random seed analysis}
% \tristan{I'm wondering if we could remove this sub-section, as it doesn't seem to be critical to the understanding of the paper, wdyt?}
% Uncertainty analysis (UA) is a method to estimate the output distribution given an input distribution.
% Sensitivity analysis (SA) is a variation of UA that measures the local response to a small perturbation~\cite{loucks2017water}. 
% Several approaches for UA and SA exist, among them differential analysis, response surface methodology, Monte Carlo analysis, and variance decomposition (see~\cite{helton2006survey} for a review). In particular, Monte Carlo analysis is a sampling-based technique that samples inputs space to estimate output response. Monte Carlo analysis shares similarities with Monte Carlo Arithmetic. For instance, the PRECISE framework~\cite{chaitin1996lectures} shares with MCA the idea of varying virtual precision to estimate output sensitivity.

% Furthermore, the quality of a Random Number Generator is a crucial characteristic when dealing with stochastic model and differ from cryptography requirements. A good RNG provides a tradeoff between performance, period, correlation, or uniformity~\cite{james2020review} for numerical simulation.  
% On the other hand, poor-quality RNGs may have a substantial impact on results~\cite{click2011quality}. First, measuring variation induced by seed is a good indicator of program stability that we can compare against variability induced by floating-point arithmetic.

@@

Visualization covers a wide range of numerical simulation domains, from general-purpose numerical visualization with tools
such as VTK~\cite{schroeder2000visualizing} to topological data analysis~\cite{tierny2018topological}, including matrix~\cite{wu2008matrix} or tensor~\cite{kindlmann2006diffusion} visualization.
We review here tools dedicated to the analysis of numerical variability induced by the floating-point model. Despite the large number of existing tools  to locate unstable source code sections, few of them allow the user to have a global view of the numerical accuracy of their computations or compare several executions of the same program under various perturbations.

Veritracer~\cite{chatelain2018veritracer} was the first tool to propose an embedded visualizer of numerical stability over time for C, C++, and Fortran codes. It is an extension of Verificarlo~\cite{verificarlo} that offers contextual information about variables traced to allow the user to identify instabilities quickly.
However, it neither offers a dynamic interface nor includes information about high-level structures such as \texttt{struct} or \texttt{array}.
FPAvizual~\cite{gu2014fpavisual} visualizes the effects of floating-point arithmetic but only for educational purposes.
Anteater~\cite{faust2019anteater} proposes a visualizer for Python codes, but as mentioned previously, it only deals with native Python types, and the user needs to select variables to trace manually. However, Anteater provides an insightful reflection 
% \tristan{It's a bit vague. Do you simply mean: ``Anteater's GUI inspired the design of \pytracer's interface"?} 
around visual debugging beyond floating-point analysis, which has inspired the design of \pytracer's visualization interface.

@@

% \subsection{Pytracer workflow}
\pytracer adopts the three-step workflow shown in Figure~\ref{fig:workflow}. First, it uses monkey patching to replace the functions called by the application with a wrapper that saves their arguments and returned values in a trace file.  Second, it injects noise in the application, executes it multiple times, and aggregates the resulting trace files to compute summary statistics about each traced variable. Finally, a Plotly dashboard provides interactive visualizations highlighting unstable code sections.

@@

\pytracer dynamically instruments Python modules
without requiring source code modifications in the application.
To do so, it creates a new, instrumented version of each module that preserves its original attributes so that the application can transparently use it.
By default, \pytracer instruments all the attributes
in a given module except the special ones of the form \texttt{\_\_*\_\_}.
However, some attributes are not writable. In this case, \pytracer preservers the original attributes and warns the user. The user can also restrict the list of traced attributes through
an inclusion/exclusion mechanism. The main advantage of this instrumentation approach is the lack of required modification in the application source code, in contrast with decorator-based approaches. In addition, this technique is scalable and applicable in read-only environments such as Singularity containers or HPC clusters.

@@

Python loads modules through an import mechanism triggered by the \texttt{import} that involves two objects: \texttt{finders} and \texttt{loaders}. \texttt{Finders} search for packages (or 
namespaces\footnote{Python distinguishes between \texttt{regular} and \texttt{namespace} packages.
A regular package is a directory that contains a \_\_init\_\_.py file and potentially subdirectories (sub-packages) 
that contain themselves a \texttt{\_\_init\_\_.py} file and so on recursively. 
The package hierarchy follows those of the directory. 
Namespace packages introduced in Python 3.3 (\href{https://www.python.org/dev/peps/pep-0420/}{PEP 420}) do not contain an
\_\_init\_\_.py file and allow for flexible directory structures. Hence, parts of the package can be located in zip files, on the network, or in separated directories. There are no functional differences between both types of packages.}) on the local storage --- such as standard packages installed through the pip package manager --- or on a distant server.
\texttt{Finders} do not load modules. Instead, they return a specification (\texttt{spec} object) that includes 
information on where to find and how to load modules.
\texttt{Loaders} create module objects, initialize them, and fill import-related attributes 
such as \texttt{\_\_spec\_\_}. 
\texttt{Loaders} then execute modules and populate their namespace. Finally, modules are bound to their import name in the \texttt{sys.modules} dictionary.

Python supports custom \texttt{finder} classes registered in the \texttt{sys.meta\_path} list.
When Python encounters an \texttt{import} statement, it first looks for an existing binding in \texttt{sys.modules} and then iterates over the \texttt{finder} classes in the \texttt{sys.meta\_path} list until it finds the module. \pytracer adds a custom \texttt{finder} class at the head of \texttt{sys.meta\_path} that intercepts
the module import and creates the module with a custom \texttt{loader} class.

\pytracer's \texttt{loader} class first loads the original module, then copies it as a new instance of the \texttt{ModuleType} class. It then calls the appropriate instrumentation function for each attribute depending on its type (function, class, or instances --- see below \tristan{iterators?}) and replaces the original module in the \texttt{sys.modules} map with the instrumented version. Finally, it updates all existing references to the original module, such as those in \texttt{\_\_globals\_\_} function attributes containing references to internal function objects \tristan{this sentence doesn't make sense, please update}. As a result, the application will transparently call the instrumented module.

@@

Listings~\ref{fig:wrapper_creation} and~\ref{fig:generic_wrapper} show \pytracer's function instrumentation mechanism.
For each function to instrument, \pytracer's \texttt{loader} class creates a wrapper function, dynamically compiles it with the compile builtin function, and substitutes it for the original module function.  However, this simple technique does not apply to callable class instances, i.e., class instances that have a \texttt{\_\_call\_\_} attribute and are not of type \texttt{FunctionType}. Indeed, applying the wrapper technique would 1) modify the type of the class instance to \texttt{FunctionType}, which could cause syntax and semantic errors, and 2) mask class attributes, leading to \texttt{AttributeError} exceptions.
\pytracer overrides the \texttt{\_\_call\_\_} attribute with the wrapper function when possible to overcome this issue.
When the \texttt{\_\_call\_\_} attribute is not writable,  Pytracer does not instrument the class and returns a warning.

@@

    Returns:
        wrapper_code: The code source of the wrapper
    """
    function_id = id(function)
    info = (function_id, function_module, function_qualified_name)
    wrapper_code = f"""
    def {function_wrapper_name}(*args, **kwargs):
        return generic_wrapper({info},*args,**kwargs)"""
    return wrapper_code
\end{lstlisting}
    \caption{Function to create the instrumented version of a function.
    \pytracer uses the identifier of the function instead of the 
    actual function to handle aliases and avoid duplicate instrumentation.}
    \label{fig:wrapper_creation}
\end{listing}

@@

\pytracer instruments classes by wrapping their callable attributes as described previously. 
This instrumentation preserves class types, which avoids type mismatches in the instrumented application.
However, classes that describe particular base types and NumPy's universal functions (class \texttt{ufunc}) contain read-only attributes that cannot be instrumented.
By default, \pytracer returns a warning when it encounters such classes.

\subsubsection{Instrumenting class instances}

\pytracer automatically instruments class instances since it instruments classes. However, it cannot instrument or instantiate classes that have read-only attributes --- in particular class \texttt{ufunc} ---, which is problematic given their pervasiveness in scientific computing. In particular, NumPy's universal functions include
vectorized elementary mathematical functions and other prevalent ones. Universal functions are implemented in C, which prevents their modification.

To overcome this issue, \pytracer wraps the class instance into a transparent wrap class (\texttt{twc}) that overloads
1) function \texttt{\_\_getattribute\_\_}, called when an instance attribute is accessed --- instrumented to return the attributes from the original class, and 2) function \texttt{\_\_call\_\_}, called when the function is called through the () operator --- instrumented with function \texttt{generic\_wrapper} . The \texttt{twc} returns the queried attribute of the instance instead of returning its attributes, allowing transparent access to the instance.  Finally, \pytracer overloads the builtin \texttt{type} function to preserve the type of the wrapped instance.
